---
title: "Count Data Analysis with GLM"
author: "Data Analyst"
date: "April 30, 2025"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
# Set global chunk options
# Hide messages and warnings by default, but display code and results
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This R Markdown document analyzes count data from an ecological experiment. We'll evaluate the effect of different treatments on species counts across time using generalized linear models (GLMs), which are appropriate for count data.

The dataset contains the following variables:
- **Block**: Experimental block (spatial replication)
- **Treatment**: Experimental treatment applied
- **Date**: Date of observation
- **species**: Species identifier
- **count**: Number of individuals counted (response variable)

## Data Preparation

First, we'll load the necessary packages and import the data.

```{r load-libraries}
# Load necessary libraries for data analysis and statistical modeling
# These packages provide essential functions for working with and analyzing the dataset
library(tidyverse)  # For data manipulation and visualization
library(lme4)       # For mixed effects models
library(car)        # For Anova function 
library(emmeans)    # For estimated marginal means (post-hoc tests)
```

```{r load-data}
# Read the data file into R and store it in an object called 'data'
# You'll need to replace this with the actual path to your data file
data <- read.csv("your_data.csv")

# Examine the structure and summary statistics of the dataset
# This helps understand variable types and distributions before modeling
str(data)
summary(data)
```

```{r data-prep}
# Convert the Date column to R's Date format for proper time-based analysis
data$Date <- as.Date(data$Date)

# Convert categorical variables to factors
# R needs to know these are categorical for proper model specification
data$Block <- as.factor(data$Block)
data$Treatment <- as.factor(data$Treatment)
data$species <- as.factor(data$species)
```

## Exploratory Data Analysis

Before building models, it's important to visualize the data to understand patterns and relationships.

```{r time-plot}
# Create a visualization of counts over time by treatment
# This helps identify temporal patterns and treatment differences
ggplot(data, aes(x = Date, y = count, color = Treatment)) +
  geom_point() +  # Plot individual data points
  geom_smooth(method = "loess") +  # Add smoothed trendlines for each treatment
  theme_bw() +  # Use a clean black and white theme for clarity
  labs(title = "Count over time by treatment",
       x = "Date", 
       y = "Count")
```

```{r species-plot}
# Create a boxplot to visualize species distribution by treatment
# This shows how counts vary across species and treatments
ggplot(data, aes(x = species, y = count, fill = Treatment)) +
  geom_boxplot() +  # Create boxplots to show distributions
  theme_bw() +  # Use clean black and white theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for better readability
  labs(title = "Count by species and treatment",
       x = "Species", 
       y = "Count")
```

## Model Fitting

Count data typically follows a Poisson distribution, but ecological count data often shows overdispersion (variance greater than mean). We'll first fit a Poisson model, then check for overdispersion, and switch to a negative binomial model if needed.

```{r poisson-model}
# Fit a Poisson GLM as the initial model for count data
# Poisson distribution is commonly used for count data
model_poisson <- glm(count ~ Treatment * Date * species, 
                    family = poisson(link = "log"), 
                    data = data)

# Check for overdispersion, which is common in ecological count data
# Overdispersion occurs when variance exceeds the mean
dispersion_stat <- sum(residuals(model_poisson, type = "pearson")^2) / df.residual(model_poisson)
print(paste("Dispersion statistic:", round(dispersion_stat, 2)))
```

```{r final-model}
# Conditional section to handle overdispersion if present
# If dispersion > 1.2, use negative binomial instead of Poisson
if(dispersion_stat > 1.2) {
  # Load MASS package which contains the negative binomial function
  # This package provides functions for more complex probability distributions
  library(MASS)
  
  # Fit negative binomial model which can handle overdispersed count data
  # This model adds a parameter to account for extra variance
  model_nb <- glm.nb(count ~ Treatment * Date * species, data = data)
  
  # Set the negative binomial model as our final model
  # This will be used for inference if overdispersion is detected
  final_model <- model_nb
  cat("Using negative binomial model due to overdispersion\n")
} else {
  # If no significant overdispersion, continue with the Poisson model
  # In this case, the simpler Poisson model is adequate
  final_model <- model_poisson
  cat("Using Poisson model (no significant overdispersion)\n")
}

# Display detailed model results
# This shows parameter estimates, standard errors, and p-values
summary(final_model)
```

## Statistical Tests

Now we'll conduct formal statistical tests to evaluate the significance of our predictors.

```{r anova}
# Conduct Type III ANOVA to test significance of main effects and interactions
# Type III tests each effect after controlling for all other effects
Anova(final_model, type = "III")
```

## Mixed Effects Model

If the blocks in the experiment represent random sampling units, we should use a mixed effects model with Block as a random effect.

```{r mixed-model, eval=TRUE}
# Fit a mixed-effects model if Block should be treated as a random effect
# This accounts for non-independence of observations within blocks
mixed_model <- glmer(count ~ Treatment * Date * species + (1|Block), 
                    family = poisson, 
                    data = data)

# Check if the mixed model converged properly
# Convergence issues are common in complex mixed models
if(is.null(mixed_model@optinfo$conv$lme4$messages)) {
  # If no convergence warnings, proceed with mixed model analysis
  # This indicates the model fitting was successful
  cat("Mixed effects model converged successfully\n")
  
  # Display detailed results of the mixed effects model
  # Shows fixed effects, random effects, and other parameters
  summary(mixed_model)
  
  # Conduct Type III ANOVA for the mixed model
  # Tests significance of effects in the mixed model framework
  car::Anova(mixed_model, type = "III")
} else {
  # If convergence issues occurred, fall back to the fixed effects model
  # This ensures we can still make valid inferences
  cat("Mixed effects model had convergence issues. Using fixed effects model instead.\n")
}
```

## Post-hoc Analysis

If significant effects are found, we need to conduct post-hoc tests to understand specific differences between treatment levels.

```{r posthoc}
# Conduct post-hoc tests to compare treatment levels
# These provide pairwise comparisons with appropriate p-value adjustments
emmeans_result <- emmeans(final_model, pairwise ~ Treatment)
print(emmeans_result)

# Calculate estimated marginal means for the Treatment by species interaction
# This shows how treatment effects might vary across species
emmeans_interaction <- emmeans(final_model, ~ Treatment | species)
print(emmeans_interaction)
```

## Model Predictions

Visualizing model predictions helps interpret complex interactions between predictors.

```{r predictions}
# Create a grid of values for prediction from the model
# This prepares data for visualizing model predictions
pred_data <- expand.grid(
  Treatment = unique(data$Treatment),  # Include all treatment levels
  Date = seq(min(data$Date), max(data$Date), length.out = 30),  # Create a sequence of dates
  species = unique(data$species)  # Include all species
)

# Add Block variable to prediction data
# Using the first block level for consistency in predictions
pred_data$Block <- levels(data$Block)[1]

# Generate model predictions on the response scale (counts)
# Transform log predictions back to the original count scale
pred_data$predicted <- predict(final_model, newdata = pred_data, type = "response")

# Create a visualization of model predictions over time
# This shows expected counts for each treatment-species combination
ggplot(pred_data, aes(x = Date, y = predicted, color = Treatment)) +
  geom_line() +  # Connect predictions with lines
  facet_wrap(~ species, scales = "free_y") +  # Create separate panels for each species
  theme_bw() +  # Use clean black and white theme
  labs(title = "Predicted counts over time by treatment and species",
       x = "Date", 
       y = "Predicted count")
```

## Model Diagnostics

Checking model assumptions is crucial for valid inference.

```{r diagnostics}
# Check model diagnostics with standard diagnostic plots
# These plots help assess if model assumptions are met
par(mfrow = c(2, 2))  # Set up a 2x2 grid of plots
plot(final_model)  # Create diagnostic plots for the final model
```

```{r model-fit}
# Define a function to extract model fit statistics
# This collects key metrics to evaluate model performance
get_model_fit <- function(model) {
  list(
    AIC = AIC(model),  # Akaike Information Criterion - lower is better
    BIC = BIC(model),  # Bayesian Information Criterion - lower is better
    deviance = deviance(model),  # Model deviance - measure of fit
    df.residual = df.residual(model)  # Degrees of freedom remaining
  )
}

# Print the model fit statistics for the final model
# These help assess the overall quality of the model
print(get_model_fit(final_model))
```

## Conclusion

This analysis evaluated the effect of Treatment on species counts across time using generalized linear models appropriate for count data. The main findings include:

1. [Insert key findings after running the analysis with your actual data]
2. [Comment on which treatments had significant effects]
3. [Describe any temporal patterns observed]
4. [Note differences in treatment effects across species if present]

## References

- Zuur, A. F., Ieno, E. N., Walker, N. J., Saveliev, A. A., & Smith, G. M. (2009). Mixed effects models and extensions in ecology with R. Springer.
- Ver Hoef, J. M., & Boveng, P. L. (2007). Quasi-Poisson vs. negative binomial regression: how should we model overdispersed count data? Ecology, 88(11), 2766-2772.