---
title: "Count Data Analysis with GLM"
author: "Diony Gamoso"
date: "April 30, 2025"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
# Set global chunk options
# Hide messages and warnings by default, but display code and results
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

This R Markdown document analyzes Count data from an ecological experiment. We'll evaluate the effect of different treatments on species Counts across time using generalized linear models (GLMs), which are appropriate for Count data.

The dataset contains the following variables:
- **Block**: Experimental block (spatial replication)
- **Treatment**: Experimental treatment applied
- **Date**: Date of observation
- **species**: Species identifier
- **Count**: Number of individuals Counted (response variable)

## Data Preparation

First, we'll load the necessary packages and import the data.

```{r load-libraries}
# Load necessary libraries for data analysis and statistical modeling
# These packages provide essential functions for working with and analyzing the dataset
library(tidyverse)  # For data manipulation and visualization
library(lme4)       # For mixed effects models
library(car)        # For Anova function 
library(emmeans)    # For estimated marginal means (post-hoc tests)
library(readxl)     # For reading Excel files
library(dplyr)      # For data manipulation and joins

```

```{r load-data}

# Read in the datasets
insect_data <- read.csv("./data/raw/2024InsectData.csv")
flower_data <- read_excel("./data/raw/2024FlowerData.xlsx")
specimen_data <- read_excel("./data/raw/2024Specimens_Prelim_ID.xlsx")

# Examine the structure of each dataset to confirm column names
str(insect_data)
str(flower_data)
str(specimen_data)

# Join insect data with flower data using DateBlkTrtmnt key
insect_flower_joined <- insect_data %>%
  left_join(flower_data, by = "DateBlkTrtmnt")

# Join insect data with specimen data (Concat_DBT_sp_num matches DG_Spec_ID_code)
insect_specimen_joined <- insect_data %>%
  left_join(specimen_data, by = c("Concat_DBT_sp_num" = "DG_Spec_ID_code"))

# Create a comprehensive dataset joining all three tables
# First join insect and flower data, then join with specimen data
all_data_joined <- insect_data %>%
  left_join(flower_data, by = "DateBlkTrtmnt") %>%
  left_join(specimen_data, by = c("Concat_DBT_sp_num" = "DG_Spec_ID_code"))

#  Save the joined datasets
write_csv(insect_flower_joined, "insect_flower_joined.csv")
write_csv(insect_specimen_joined, "insect_specimen_joined.csv")
write_csv(all_data_joined, "all_data_joined.csv")


data<-insect_specimen_joined %>%
    select(Date, Block, Treatment, CountType, Insect.type, Jaime_Genus_ID.y, Jaime_Species_ID.y, Jaime_GenusSpeciesID.y, Jaime_Sex_ID.y, Count) %>% # Select relevant columns
  filter(!is.na(Count)) %>%  # Remove rows with missing Counts
  # Filter out rows where CountType is "snapshot"
  filter(CountType != "snapshot") %>%
  mutate(Date = as.Date(Date, format = "%d%b%Y"))  # Convert Date to Date type
  
```

```{r data-prep}

# Convert categorical variables to factors
# R needs to know these are categorical for proper model specification
data$Block <- as.factor(data$Block)
data$Treatment <- as.factor(data$Treatment)
#data$species <- as.factor(data$species)
```

## Exploratory Data Analysis

Before building models, it's important to visualize the data to understand patterns and relationships.

```{r time-plot}
# Create a visualization of Counts over time by treatment
# This helps identify temporal patterns and treatment differences
ggplot(data, aes(x = Date, y = Count, color = Treatment)) +
  geom_point() +  # Plot individual data points
  geom_smooth(method = "loess") +  # Add smoothed trendlines for each treatment
  theme_bw() +  # Use a clean black and white theme for clarity
  labs(title = "Count over time by treatment",
       x = "Date", 
       y = "Count")
```


```

## Model Fitting

Count data typically follows a Poisson distribution, but ecological Count data often shows overdispersion (variance greater than mean). We'll first fit a Poisson model, then check for overdispersion, and switch to a negative binomial model if needed.

```{r poisson-model}
# Fit a Poisson GLM as the initial model for Count data
# Poisson distribution is commonly used for Count data, include Block as a Fixed Effect
model_poisson_block_fixed <- glm(Count ~ Treatment * Date + Block , # Include interaction between Treatment and Date
                    family = poisson(link = "log"), 
                    data = data)

# Check for overdispersion, which is common in ecological Count data
# Overdispersion occurs when variance exceeds the mean, if dispersion > 1.2, use negative binomial instead of Poisson
dispersion_stat <- sum(residuals(model_poisson, type = "pearson")^2) / df.residual(model_poisson)
print(paste("Dispersion statistic:", round(dispersion_stat, 2)))

# Summary of the model
summary(model_poisson_block_fixed)

# Model 2: Include Block as an interaction with Treatment
# This allows for treatment effects to vary across blocks
model_poisson_block_interaction <- glm(Count ~ Treatment * Date + Treatment * Block, 
                                      family = poisson(link = "log"), 
                                      data = data)

# Summary of the model with block interaction
summary(model_poisson_block_interaction)
  
```
##Model results

Main Effects


Treatment effects:

Both "h" treatment (145.4) and "nh" treatment (143.6) show significant positive effects (p < 0.05), indicating higher insect counts compared to the control treatment.


Date effect:

The significant positive Date coefficient (0.006892) suggests that insect counts generally increase over time across all treatments.


Block effects:

Several blocks show significant differences from the reference block (likely "fsgc" or another block that doesn't appear as a significant term):
Blocks "fosc", "lile", "poof", "wewa", and "wywe" all have significant negative coefficients, indicating lower insect counts in these blocks compared to the reference block.



Interaction Effects

Treatment × Date interactions:

Both "h" (-0.0073) and "nh" (-0.00718) interactions are significant and negative.
This means that while counts increase over time in the control group, the rate of increase is significantly lower in the "h" and "nh" treatment groups (the date effect is partially canceled out).


Treatment × Block interactions:

Multiple significant interactions, particularly for "h" treatment, suggest that treatment effects vary across blocks.
For example, the "h" interaction (2.704) is positive and significant, indicating that the "h" treatment effect is stronger in the "fosc" block compared to the reference block.
There are "NA" values for some interactions, indicating singularities (perfect collinearity), likely because the reference block and treatment combination contains no observations.



Model Fit

The residual deviance (230.08) is much lower than the null deviance (342.34), suggesting the model explains a substantial portion of the variation.
The AIC is 1635.1, which can be used to compare this model with alternative models.
78 observations were deleted due to missingness, which might affect your results.

Key Takeaways

Both treatments ("h" and "nh") significantly increase insect counts compared to the control.
There's a general trend of increasing counts over time.
The effect of treatments diminishes over time (negative Treatment interactions).
There's substantial spatial variation (block effects), and treatment effectiveness varies by location (Treatment interactions).
The model doesn't show signs of severe overdispersion since the dispersion parameter is set to 1 for the Poisson family.

To improve your analysis, you might consider:

Checking for overdispersion more explicitly
Examining treatment effects at specific time points of interest
Visualizing the interaction effects to better understand how treatments perform across different blocks
## Statistical Tests

Now we'll conduct formal statistical tests to evaluate the significance of our predictors.

```{r anova}
# Conduct Type III ANOVA to test significance of main effects and interactions
# Type III tests each effect after controlling for all other effects
Anova(model_poisson_block_interacton, type = "III")
```

## Mixed Effects Model

If the blocks in the experiment represent random sampling units, we should use a mixed effects model with Block as a random effect.

```{r mixed-model, eval=TRUE}
# Fit a mixed-effects model if Block should be treated as a random effect
# This acCounts for non-independence of observations within blocks
mixed_model <- glmer(Count ~ Treatment * Date  + (1|Block), 
                    family = poisson, 
                    data = data)

# Check if the mixed model converged properly
# Convergence issues are common in complex mixed models
if(is.null(mixed_model@optinfo$conv$lme4$messages)) {
  # If no convergence warnings, proceed with mixed model analysis
  # This indicates the model fitting was successful
  cat("Mixed effects model converged successfully\n")
  
  # Display detailed results of the mixed effects model
  # Shows fixed effects, random effects, and other parameters
  summary(mixed_model)
  
  # Conduct Type III ANOVA for the mixed model
  # Tests significance of effects in the mixed model framework
  car::Anova(mixed_model, type = "III")
} else {
  # If convergence issues occurred, fall back to the fixed effects model
  # This ensures we can still make valid inferences
  cat("Mixed effects model had convergence issues. Using fixed effects model instead.\n")
}
```

## Post-hoc Analysis

If significant effects are found, we need to conduct post-hoc tests to understand specific differences between treatment levels.

```{r posthoc}
# Conduct post-hoc tests to compare treatment levels
# These provide pairwise comparisons with appropriate p-value adjustments
emmeans_result <- emmeans(final_model, pairwise ~ Treatment)
print(emmeans_result)


```

## Model Predictions

Visualizing model predictions helps interpret complex interactions between predictors.

```{r predictions}
# Create a grid of values for prediction from the model
# This prepares data for visualizing model predictions
pred_data <- expand.grid(
  Treatment = unique(data$Treatment),  # Include all treatment levels
  Date = seq(min(data$Date), max(data$Date), length.out = 30),  # Create a sequence of dates
  species = unique(data$species)  # Include all species
)

# Add Block variable to prediction data
# Using the first block level for consistency in predictions
pred_data$Block <- levels(data$Block)[1]

# Generate model predictions on the response scale (Counts)
# Transform log predictions back to the original Count scale
pred_data$predicted <- predict(final_model, newdata = pred_data, type = "response")

# Create a visualization of model predictions over time
# This shows expected Counts for each treatment-species combination
ggplot(pred_data, aes(x = Date, y = predicted, color = Treatment)) +
  geom_line() +  # Connect predictions with lines
  facet_wrap(~ species, scales = "free_y") +  # Create separate panels for each species
  theme_bw() +  # Use clean black and white theme
  labs(title = "Predicted Counts over time by treatment and species",
       x = "Date", 
       y = "Predicted Count")
```

## Model Diagnostics

Checking model assumptions is crucial for valid inference.

```{r diagnostics}
# Check model diagnostics with standard diagnostic plots
# These plots help assess if model assumptions are met
par(mfrow = c(2, 2))  # Set up a 2x2 grid of plots
plot(final_model)  # Create diagnostic plots for the final model
```

```{r model-fit}
# Define a function to extract model fit statistics
# This collects key metrics to evaluate model performance
get_model_fit <- function(model) {
  list(
    AIC = AIC(model),  # Akaike Information Criterion - lower is better
    BIC = BIC(model),  # Bayesian Information Criterion - lower is better
    deviance = deviance(model),  # Model deviance - measure of fit
    df.residual = df.residual(model)  # Degrees of freedom remaining
  )
}

# Print the model fit statistics for the final model
# These help assess the overall quality of the model
print(get_model_fit(final_model))
```

## Conclusion

This analysis evaluated the effect of Treatment on species Counts across time using generalized linear models appropriate for Count data. The main findings include:

1. [Insert key findings after running the analysis with your actual data]
2. [Comment on which treatments had significant effects]
3. [Describe any temporal patterns observed]
4. [Note differences in treatment effects across species if present]

## References

- Zuur, A. F., Ieno, E. N., Walker, N. J., Saveliev, A. A., & Smith, G. M. (2009). Mixed effects models and extensions in ecology with R. Springer.
- Ver Hoef, J. M., & Boveng, P. L. (2007). Quasi-Poisson vs. negative binomial regression: how should we model overdispersed Count data? Ecology, 88(11), 2766-2772.